---
title: "Assignment 6: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Joshua Frear"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A06_GLMs.Rmd") prior to submission.

The completed exercise is due on Tuesday, March 2 at 1:00 pm.

## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup}
#1
getwd()
library(tidyverse)
library(agricolae)
library(lubridate)
#import data
NTL.chem.physics.raw <- read.csv("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv")
str(NTL.chem.physics.raw)
#set sample date to Date vector type
NTL.chem.physics.raw$sampledate <- as.Date(NTL.chem.physics.raw$sampledate, "%m/%d/%y")

#2
mytheme <- theme_gray(base_size = 12) +
  theme(legend.position = "top") 

theme_set(mytheme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: Mean lake temperatures in July at different depths, across all lakes, are not different.
Ha: Mean lake temperatures in July at different depths, across all lakes, are different.

4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
NTL.July.temps <-
  NTL.chem.physics.raw %>% 
  filter(month(sampledate) == 7) %>% 
  select(lakename, year4, daynum, depth, temperature_C) %>% 
  filter(!is.na(temperature_C) & !is.na(depth)) 

#5
plot1 <- ggplot(NTL.July.temps, aes(x = jitter(depth), y = temperature_C)) + geom_point() + geom_smooth(method = "lm", color = "red", size = 1.7) + ylim(0, 35) + xlab("Depth") + ylab(expression(paste("Temperature, ", ~degree, "C")))
print(plot1)

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: It appears that temperature descreases with increasing depth, at least until a depth of 7 or 8, after which temperature does not respond to changes in depth. The line corresponds to roughly the midpoint of temperature ranges from depths 0 until 7 or 8, and then after that, it fits the data extremely poorly, as temperatures level out. This suggests that the trend is not linear. 


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7
lake.temp.model <- lm(data = NTL.July.temps, temperature_C ~ depth)
summary(lake.temp.model)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: According to the linear model, temperature measurements decrease by about 1.95 degrees C for every meter deeper in the lake. The correlation of the variables is statistically significant (p < 0.001). This model uses depth to explain 73.87% of the variability in temperature, based on 9726 degrees of freedom. 

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
full.model <- lm(data = NTL.July.temps, formula = temperature_C ~ year4 + daynum + depth)
step(full.model)

#10
full.model <- lm(data = NTL.July.temps, formula = temperature_C ~ year4 + daynum + depth)
summary(full.model)
```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The AIC method returns the full set of variables as the best model (year4, daynum, and depth). This model explains 74.11% of the variance in temperature (adjusted R-squared). This is an improvement over the model that uses only depth (73.87% explained) but extremely marginally so. Because the model using only depth is so parsimonious and the models have very similar R-squared values, I would actually prefer the model with fewer variables.

---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
lake.temp.aov <- aov(data = NTL.July.temps, formula = temperature_C ~ lakename)
summary(lake.temp.aov)

lake.temp.lm <- lm(data = NTL.July.temps, formula = temperature_C ~ lakename)
summary(lake.temp.lm)
```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Yes, the differences in mean temperature are significant (p < 0.005).

14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2, fig.align = "center"}
#14.
library(RColorBrewer)
plot2 <- ggplot(NTL.July.temps, aes(x = jitter(depth), y = temperature_C, color = lakename)) + geom_point(alpha = 0.5) + scale_color_brewer(palette="Paired") + geom_smooth(method = "lm", se = FALSE) + ylim(0,35) + xlab("Depth (m)") + ylab(expression(paste("Temperature (", ~degree, "C)"))) +  scale_x_continuous(expand = c(0, 0)) + theme(legend.position = "bottom") + labs(color = "Lakes") + theme(legend.text=element_text(size=8), legend.key.size = unit(8, "pt"))
print(plot2)
```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(lake.temp.aov)
#unique(NTL.July.temps$lakename)
```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Ward Lake and Paul Lake have the same mean temperature as Peter Lake, according to the TukeyHSD() test. No lake has a mean temperature that is statistically distinct from all other lakes.

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: a t-test or f-test.